Arriaga meets Kitagawa. Life expectancy decomposition and population subgroups

### 2024-09-12 manuscript/manuscript.tex

%Version 2.1 April 2023
% See section 11 of the User Manual for version history
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[sn-apa,pdflatex]{sn-jnl}

@@

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published
%%%%  by Springer Nature. The guidance has been prepared in partnership with
%%%%  production teams to conform to Springer Nature technical requirements.
%%%%  Editorial and presentation requirements differ among journal portfolios and
%%%%  research disciplines. You may find sections in this template are irrelevant
%%%%  to your work and are empowered to omit any such section if allowed by the
%%%%  journal you intend to submit to. The submission guidelines and policies
%%%%  of the journal take precedence. A detailed User Manual is available in the
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

%% Per the spinger doc, new theorem styles can be included using built in style, 
%% but it seems the don't work so commented below
%\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

%% \theoremstyle{thmstyletwo}%
\theoremstyle{remark}
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

%% \theoremstyle{thmstylethree}%
\theoremstyle{definition}
\newtheorem{definition}{Definition}%
\usepackage{amsmath}

\raggedbottom

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\title[Arriaga meets Kitagawa]{Arriaga meets Kitagawa. life expectancy
decompositions including population subgroups}

%%=============================================================%%
%% Prefix	-> \pfx{Dr}
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% NatureName	-> \tanm{Poet Laureate} -> Title after name
%% Degrees	-> \dgr{MSc, PhD}
%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate}
%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
%%=============================================================%%

@@

  \affil*[1]{\orgdiv{Department of Sociology and Social
Work}, \orgname{University of the Basque Country
(UPV/EHU)}, \orgaddress{\city{Leioa}, \country{Spain}, \postcode{48940}, \state{Bizkaia}, \street{Barrio
Sarriena s/n}}}
  \affil[2]{\orgname{Ikerbasque (Basque Foundation for Science)}}
  \affil[3]{\orgdiv{Laboratory of Population Health}, \orgname{Max
Planck Institute for Demographic Research}}

\abstract{\textbf{Background}: An Arriaga (1984) decomposition allows us
to decompose differences in life expectancy into contributions due to
mortality rate differences in each age. A Kitawaga (1955) decomposition
allows us to decompose differences in a weighted mean into an effect
from differences in structure and effects from differences in each
element of the weighted value.

\textbf{Objective:} Sometimes we would like to decompose a difference
between two populations that are each composed of like-defined
subpopulations. Said decomposition would produce effects for each age of
each subpopulation, as well as a marginal effect of composition
differences. I propose a straightforward analytic method to do this.

\textbf{Methods:} In short, within-subpopulation life expectancy
differences can be handled by the Arriaga method. The case of weighting
together the life expectancies of subpopulations by way of a composed
radix can be handled using the well-known Kitagawa method. The elements
of the value component of the Kitagawa method tell us how to rescale the
Arriaga results specific to each subpopulation.

\textbf{Results:} I show the proposed analytic method to be equivalent
to a Horiuchi et al (2008) reframing of the same problem. Some
mentionable properties: (i) There is no limit to the number of
subpopulations, (ii) it is straightforward to incorporate cause-of-death
information, (iii) composition is here only considered in the radix age.
I currently have results for simulated mortality rates, but I promise to
wrangle up an empirical application to demonstrate the method.

\textbf{Conclusions:} The analytic decomposition method I propose is
advantageous compared to a Horiuchi method for this problem purely for
reasons of computational efficiency. This method could help further
disentangle the effects of mortality and composition differences in
explaining or clarifying paradoxes or secular change. I promise to think
further about subpopulation weighting that might occur over all ages
(using prevalence information).\}}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

An \citet{arriaga1984measuring} decomposition allows us to decompose
changes in life expectancy into contributions due to mortality changes
in different ages. The method was designed to be practical, and framed
in terms of age lifetable columns expressed in discrete age groups. A
well-known property of the method is that mortality changes in different
ages need not be proportional. Derived contributions also sum exactly to
the observed life expectancy difference. A not-well-known property of
the method is that it is asymmetrical, in the sense that the absolute
values of age-specific contributions depend on whether we compare
population 1 with 2, or population 2 with 1. A further property of the
method is that it is designed to work with homogeneous populations,
meaning that populations 1 and 2 are each homogeneous, in the sense that
each is expressed with only one lifetable.

A \citet{kitagawa1955components} decomposition allows us to decompose
differences in a weighted mean due to differences in weights (structure)
and differences in the value being weighted (often rates, but in our
case life expectancies). This widely-used decomposition method is
well-known to be exact in that the resulting structure and value
components sum exactly to the observed difference in weighted means. The
individual elements (ages, or life expectancies for us) of the value
being weighted have identifiable effects. It is not well-known that the
individual elements of the structure component do not have identifiable
effects. Rather, the structure effects should be summed to a a marginal
effect due to differences in structure.

Sometimes we have a situation where we would like to decompose a
difference between two populations that are each composed of
like-defined subpopulations. For example, life expectancy in France
versus Spain, each with education-specific subpopulations. In such
situations, a decomposition should tell us the contribution to the
difference in overall life expectancy due to rate differences in each
age in each subpopulation, and also separate an effect due to overall
compositional change. In this paper, I would like to propose a
straightforward analytic method to decompose in this way. In short,
within-subpopulation life expectancy differences can be handled by the
Arriaga method. The case of weighting together the life expectancies of
subpopulations by way of a composed radix can be handled using the
well-known Kitagawa method. The elements of the value component of the
Kitagawa method tell us how to rescale the Arriaga results specific to
each subpopulation so as to isolate the age-subpopulation-specific
effects on the overall life expectancy difference. I will justify this
rescaling, and show that the results of this procedure are fully
consistent with the results of a \citet{horiuchi2008decomposition}
reframing of the problem.

\hypertarget{method}{%
\section{Method}\label{method}}

\hypertarget{notation}{%
\subsection{Notation}\label{notation}}

Arriaga decompositions can be implemented in a number of relevant ways,
per Riffe et. al.~(2024), from which we here consider only
\emph{symmetrical} Arriaga decomposition. This implies following the
original formulas twice, once in each direction, and averaging the
sign-adjusted results. For this, we use the following lifetable columns:

\begin{itemize}

\item{$\ell(a)$} lifetable survivorship at exact age $a$.

\item{${}_nL_a$} lifetable person years lived in the interval $[a,a+n)$.

\item{$T_a$} total lifetable person years lived beyond age $a$.

@@

\end{itemize}

We also use the superscript \(s\) to index subpopulations comprising the
total population, and the superscript \(t\) to index time points, or
some other way of differentiating the total population. The index
\(e^{s,t}_0\) reads as ``life expectancy at birth for subpopulation
\(s\) at time \(t\)'', e.g.~\(e^{low, 1990}_0\) could be the life
expectancy in 1990 of a low education group. The first stage of our
exercise is to decompose subgroup-specific differences in life
expectancy (hence the \(^s\) on everything):

\begin{align}
{}\Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
 &= \sum _x^\omega {}_n\Delta_x^s \quad \textrm{,}\nonumber 
\end{align}

\noindent where \(\omega\) is the highest (open) age group, and
\(\Delta^s\) is the subgroup-specific (\(s\)) difference in life
expectancy being decomposed, and it is composed of age-specific
contributions, \({}_n\Delta_x^s\), which can be calculated following
Arriaga's decomposition method following eq \eqref{eq_arriaga1}. I'll
use a lifetable radix of 1, meaning \(\ell_0 = 1\) to reduce the formula
a bit.

\begin{equation}
\label{eq_arriaga1}
 {}_n\Delta_x^s = \begin{cases}\ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)  & \text{$ \forall x < \omega$}  \\
  \ell_\omega ^{s,1} \cdot (e_\omega^{s,2} -e_\omega^{s,1} ) & \text{$\forall x = \omega $}
  \end{cases}
\end{equation}

Equation \eqref{eq_arriaga1} is the first pass of our symmetrical
decomposition, where we move \emph{forward} in time from \(t=1\) to
\(t=2\), so please permit . It

>>>>

\documentclass[12pt, sn-apa,pdflatex,letterpaper]{sn-jnl}

@@

\usepackage[T1]{fontenc}
\usepackage{tgbonum}
\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
\usepackage{natbib}
\usepackage[pass]{geometry}
\usepackage{makecell}

\usepackage{lineno}
\linenumbers

%%%%
\raggedbottom
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\renewcommand{\baselinestretch}{1.5}

@@

\title[Arriaga meets Kitagawa]{Arriaga meets Kitagawa. Life expectancy decomposition and population subgroups}

@@

\author*[1]{ \fnm{Rustam} \sur{Tursun-zade} }

\author*[4]{\pfx{Dr.}  \fnm{Sergi} \sur{Trias Llimós} }

  \affil*[1]{\orgdiv{OPIK, Department of Sociology and Social
Work}, \orgname{University of the Basque Country
(UPV/EHU)}, \orgaddress{\city{Leioa}, \country{Spain}, \postcode{48940}, \state{Bizkaia}, \street{Barrio
Sarriena s/n}}}
  \affil[2]{\orgname{Ikerbasque (Basque Foundation for Science)}}
  \affil[3]{\orgdiv{Laboratory of Population Health}, \orgname{Max
Planck Institute for Demographic Research}}
\affil[4]{ \orgname{Centre for Demographic Studies}}

\doublespacing

\abstract{\textbf{Background}: An Arriaga (1984) decomposition partitions differences in life expectancy into contributions from mortality rate differences in each age. A Kitagawa (1955) decomposition partitions a difference between two weighted means into effects from differences in structure and from differences in each element of the weighted value.

\textbf{Objective:} To propose an analytic method to decompose the difference in life expectancy between two populations, each consisting of like-defined subpopulations. Said decomposition would yield effects for each age within each subpopulation and a marginal effect that accounts for composition differences. We illustrate it by examining the sex gap in life expectancy in Spain using education-specific mortality data.

\textbf{Methods:} Life expectancy differences between like-defined subpopulations can be decomposed using the Arriaga method, or a variant thereof. The results of subgroup-specific decompositions can be weighted together using the value component from a Kitagawa decomposition of total life expectancy, given that subpopulations are blended in the radix age. The composition component of the same Kitagawa decomposition gives the effect of subgroup composition on total life expectancy differences.

\textbf{Results:} Notable properties of the method include: (i) it accommodates any number of subpopulations, (ii) it easily incorporates cause-of-death information, and (iii) composition is considered in the radix age. We apply the method to Spanish cause- and education-specific data. We demonstrate that the proposed analytic method is equivalent to reframing the same problem using the generalized linear integral decomposition method.

\textbf{Conclusions:} This method can further disentangle the effects of mortality and composition differences, helping to explain or clarify paradoxes and secular change. Our analytic decomposition method is more computationally efficient than the linear integral or stepwise replacement decomposition methods. We give both R code and spreadsheet implementations of the method.}

@@

\section*{Introduction}\label{intro}

Population-level differences in life expectancy have been widely analyzed using decomposition methods to understand the impacts of mortality differences in age groups \citep[e.g.,][]{arriaga1984measuring, pollard1988decomposition},  or \citet{kitagawa1955components} to understand better the impact of compositional differences in social strata \citep{kitagawa1977mortality}. It has long been known among demographers that improvements in social welfare have the potential to bring large scale improvements in mortality conditions even net of medical improvements \citep{kitagawa1968education}. However, very few demographers have ventured to account for both age-specific effects and compositional effects at the same time \citep{torres2019contribution, hendi2021immigration, su2023educational, su2024subnational},. Disentangling the compositional effect within a conventional age group decomposition would improve our understanding of life expectancy differences between populations.

The life expectancy decomposition method proposed by \citet{arriaga1984measuring} separates a change in life expectancy into contributions due to mortality changes at different ages. This technique is designed to be practical in that it is framed in terms of lifetable columns expressed in discrete age groups. Two well-known properties of the method are that mortality changes in different ages do not need to be proportional and that the derived contributions sum exactly to the observed difference in life expectancy. A documented but lesser-known property of this method is its asymmetry: the absolute values of age-specific contributions depend on whether we compare population one with two, or vice versa. This is why the method is described in terms of changes rather than group differences; the direction of time is clear, so we always decompose an earlier period against a later period. Importantly, the method is designed to work with homogeneous populations, meaning a single lifetable represents each population. Let's call this method and the variant of it that we later apply the ``Arriaga method''. The Arriaga method has different extensions available to separate the effects of different causes of death within age groups. We seek to extend this method to account for lifetables that are composed of subpopulations with different mortality.

The decomposition method proposed by \citet{kitagawa1955components} separates differences in a weighted mean into contributions from differences in weights (structure) and the value being weighted. Often, the weighted values are rates, but in our case, the weighted values will be the life expectancies of subgroups. This method is well-known to partition differences owing to ``structure'' and ``value'' components, which sum exactly to the observed difference in weighted means. The individual elements of the value being weighted (in our case, life expectancies) have identifiable effects. It is not well-known that the individual elements of the structure component do not have identifiable effects. Instead, the structure effects should be summed up as a total marginal effect due to composition differences. We refer to this method as the ``Kitagawa'' method.

We propose to treat the problem of life expectancy decomposition by age, where a total life expectancy is calculated as the weighted mean of the expectancies of subpopulations. This is only practical when each population is composed of like-defined subpopulations. For example, we may decompose the life expectancy difference between two countries considering differences in education structure, since life expectancy varies by education level \citep{mackenbach2019determinants, trias2023age}. Or, one might wish to calculate a national life expectancy as the weighted average of its regions at two points in time since regional weightings and inequalities may have changed over time. In such scenarios, a decomposition should yield the contribution to the overall life expectancy difference stemming from rate differences in each age (and potentially causes of death) within each subpopulation while giving the overall effect of compositional differences. We propose a straightforward analytic method to perform such decompositions.

In short, differences in life expectancy between subpopulations can be decomposed using the Arriaga method, one of its variants, or by any other life expectancy decomposition method that is acceptably precise with discrete data \citep{Ponnapalli2005,horiuchi2008decomposition,andreev2002algorithm}. In a second step, the Kitagawa method is then used to rescale group-specific life expectancy decompositions, yielding an extra component telling us the importance of compositional differences. We justify this second-pass rescaling and demonstrate that the results of this procedure align with those of a linear integral reframing of the problem, per \citet{horiuchi2008decomposition}. We then apply these methods to decompose the sex gap between Spanish men and women in remaining life expectancy at age 35, considering differences in composition by education attainment around the same age and considering cause-specific mortality rates.

\section*{Method}\label{method}

\subsection*{Notation}\label{notation}

We use the following variables and scripting, most of which are lifetable columns:

\begin{itemize}
\item{${}_nm_a$} mortality rate in the age interval $[a,a+n)$.

\item{$\ell_a$} lifetable survivorship at exact age $a$.

\item{${}_nL_a$} lifetable person-years lived in the interval $[a,a+n)$.

\item{$T_a$} total lifetable person-years lived beyond age $a$.

@@

\item{$\pi_a^s$} the population fraction for subgroup $s$ in age $a$.
\end{itemize}

We use the superscript $s$ to index subpopulations comprising the total population, and the superscript ``$1$'' to typically indicate the population with the lower value of $e_0$, and ``$2$'' for the higher value, such that their difference, $\Delta = e_0^2 - e_0^1$, is always positive. Causes of death may be indicated with a $c$ subscript, i.e. $m_{a,c}$. Throughout the manuscript, we assume single-year age groups and omit age interval (left-side) subscripts ($n$) where they would otherwise be due.

\subsection*{Averaging life expectancy}\label{averaging-life-expectancy}

\citet{vaupel2002life} distinguishes between two major approaches to calculate life expectancy for a total population composed of observed subgroups. The dominant approach is based on averaging the mortality rate in each age by aggregating deaths and exposures over all subpopulations and calculating the rate per equation~\eqref{eq_rates}:

\begin{equation}
\label{eq_rates}
m_a = \frac{\sum_i{D_a^i}}{\sum_i{E_a^i}} = \sum_i \pi_a^i m_a^i \quad \mathrm{,}
\end{equation}

\noindent where $D$ and $E$ stand for deaths and exposures, respectively. This is the standard method used implicitly or explicitly by national statistical offices, the Human Mortality Database \citep{wilmoth2021methods}, or the World Population Prospects \citep{wpp2022mp}, among others. \citet{vaupel2002life} called this approach the \emph{current rates} perspective, which treats a heterogeneous population as a homogeneous one.

A second approach derives independent subgroup-specific lifetables and then weights life expectancies together based on an initial mixing composition to obtain the total life expectancy per equation~\eqref{eq_avg}:

\begin{equation}
\label{eq_avg}
e_0 = \sum_s \pi_0^s \cdot e_0^s \quad \textrm{.}
\end{equation}

\noindent This method aligns with Vaupel's \emph{current conditions} approach, at least to the extent that heterogeneous conditions are accounted for using observed strata. This second approach is commonly applied in multistate models with strata combined according to mixed initial conditions to obtain the total population \citep{caswell2018age, caswell2021healthy}, occasionally also with standard lifetables \citep{gupta1988note, muszynska2022mortality}, or when calculating between-within decompositions of variance or other similar summary measures \citep{seaman2019changing, lifeineq}. If the composition is more favorable in younger than in older ages (for example when younger generations have higher average educational attainment), the life expectancy obtained with \eqref{eq_avg} will be higher than one based on \eqref{eq_rates}. This is the total life expectancy approach that our proposed method is designed to decompose.

\subsection*{Kitagawa decomposition}\label{kitagawa-decomposition}

Equation \eqref{eq_avg} treats the total life expectancy as a weighted average, allowing us to precisely decompose a difference in $e_0$, i.e. where $\Delta = e_0^2 - e_0^1$, using the formulas from \citet{kitagawa1955components}.  We presume that population 2 has the higher life expectancy of the two, such that $\Delta$ is positive. Specifically, equation \eqref{eq_kit_comp} gives an overall effect of differences in composition:

\begin{equation}
\label{eq_kit_comp}
\Delta^{(\pi)} = \sum_s \left(\pi^{s,2}- \pi^{s,1}\right) \cdot \overline{e_0^s}\quad \textrm{,}
\end{equation}
 where 
 \begin{equation}\label{eq:e0mean}
     \overline{e_0^s} = \frac{e_0^{s,1} + e_0^{s,2}}{2} 
 \end{equation}

\noindent This result is widely known. Note that the the composition effect must be summed this way because group-specific composition (structure) effects are not well-identified and therefore cannot be interpreted in isolation. Equation \eqref{eq_kit_e0} gives the subgroup-specific effects of differences in life expectancy:

\begin{equation}
\label{eq_kit_e0}
\Delta^{(e_0^s)} = \left(e_0^{s,2} - e_0^{s,1}\right) \cdot \overline{\pi^s}\quad \textrm{,}
\end{equation}
where $\overline{\pi^s}$ is the average composition, analogous to equation~\eqref{eq:e0mean}, and where the superscripts are consistent with those used to calculate the total difference $\Delta$.

Equation~\eqref{eq_kit_sum} states that the observed difference in total life expectancy is the sum of (i) a single component capturing the effect of compositional change and (ii) a set of components giving the contribution of each subgroup's life expectancy difference to the total life expectancy difference. This second effect could equivalently be called the \textit{rate} effect because each life expectancy is ultimately a function of mortality rates.

\begin{equation}
\label{eq_kit_sum}
\Delta = \Delta^{(\pi)}  + \sum_s \Delta^{(e_0^s)}\textrm{.}
\end{equation}

\subsection*{Symmetrical Arriaga
decomposition}\label{symmetrical-arriaga-decomposition}

The \emph{rate} effect as per \eqref{eq_kit_e0} can be viewed as the \emph{net} effect on total life expectancy differences resulting from the age-specific (potentially also cause-specific) differences between like subgroups, as isolated by various life expectancy decomposition methods. In this setting, the choice of method to derive age-specific effects for subgroup-specific changes in life expectancy is not crucial. The \citet{arriaga1984measuring} decomposition approach is widely favored because it is designed for discrete data, it is analytic (implying computational efficiency), and because decomposition results sum exactly to the observed life expectancy difference. Since our application concerns sex differences rather than changes over time, we consider a \emph{symmetrical} Arriaga decomposition. Since we also include cause-of-death information, we transform the symmetrical decomposition results into an implied sensitivity function. This exercise is repeated for each subgroup to decompose group-specific changes in life expectancy (indicated with the superscript $^s$ on each variable):

\begin{equation}
\label{eq_kit_e0_s}
\Delta^s &= e_0^{s,2} - e_0^{s,1} 
\end{equation}

\noindent In this equation $\Delta^s$ is the subgroup-specific ($s$) difference in life expectancy being decomposed, which is composed of age-specific contributions, $\overrightarrow{\Delta_a^s}$ or $\overleftarrow{\Delta_a^s}$ depending on whether we decompose from population 1 to 2 (forward) or from 2 to 1 (backward). The forward age-specific values $\overrightarrow{\Delta_a^s}$ can be calculated following Arriaga's decomposition method, consistent with \citet{arriaga1984measuring} or the presentation in \citet{preston2000demography}, as outlined in equation \eqref{eq_arriaga1}. We use a lifetable radix of 1, meaning $\ell_0 = 1$, to simplify the formula slightly.

\begin{equation}
\label{eq_arriaga1}
 \overrightarrow{\Delta_a^s} = \begin{cases}\ell_a^{s,1} \cdot \left( \frac{L_a^{s,2}}{\ell_a^{s,2}} - \frac{L_a^{s,1}}{\ell_a^{s,1}}\right)+T_{a+n}^{s,2}\cdot\left(\frac{\ell_a^{s,1}}{\ell_a^{s,2}} - \frac{\ell_{a+n}^{s,1}}{\ell_{a+n}^{s,2}}\right)  & \text{$ \forall a < \omega$}\textrm{,}  \\
  \ell_\omega ^{s,1} \cdot (e_\omega^{s,2} -e_\omega^{s,1} ) & \text{$\forall a = \omega $}\textrm{.}
  \end{cases}
\end{equation}

Equation \eqref{eq_arriaga1} represents the first pass of our symmetrical decomposition, while \eqref{eq_arriaga2} is the second pass, which is identical except for swapping population superscripts.

\begin{equation}
\label{eq_arriaga2}
 \overleftarrow{\Delta_a^s} = \begin{cases}\ell_a^{s,2} \cdot \left( \frac{L_a^{s,1}}{\ell_a^{s,1}} - \frac{L_a^{s,2}}{\ell_a^{s,2}}\right)+T_{a+n}^{s,1}\cdot\left(\frac{\ell_a^{s,2}}{\ell_a^{s,1}} - \frac{\ell_{a+n}^{s,2}}{\ell_{a+n}^{s,1}}\right)  & \text{$ \forall a < \omega$}\textrm{,}  \\
  \ell_\omega ^{s,2} \cdot (e_\omega^{s,1} -e_\omega^{s,2} ) & \text{$\forall a = \omega $}\textrm{.}
  \end{cases}
\end{equation}

Importantly,

\begin{align}
\label{eq_sum2}
 -\Delta^s &= \sum_a \overleftarrow{\Delta_a^s} \\
 &= e_0^1 - e_0^2 \nonumber\textrm{.}
\end{align}

A symmetrical estimate of $\Delta_a^s$ is given by the sign-adjusted average of \eqref{eq_arriaga1} and \eqref{eq_arriaga2} for each age, again with a preference for a positive difference:

\begin{equation}
\label{eq_arriaga3}
\Delta_a^s = \frac{\left(\overrightarrow{\Delta_a^s} -  \overleftarrow{\Delta_a^s}\right)}{2}\textrm{.}
\end{equation}

Repeat the symmetrical Arriaga steps to derive age-specific contributions, $\Delta_a^s$, for each life expectancy difference between like-defined subgroups. To take into account information on causes of death, it is best to derive an ad-hoc all-cause sensitivity function, $s_a^s$, by dividing the decomposition result by the mortality rate difference:

\begin{equation}
\label{eq:arr_sen}
s_a^s = \frac{\Delta_a^s}{m_{a}^{s,2} - m_{a}^{s,1}} \quad\mathrm{,}
\end{equation}

\noindent and then multiply age-cause-specific mortality rate differences into $s_a^s$ to obtain the subgroup-specific decomposition by age and cause per equation~\eqref{eq:arr_mxc}.

\begin{equation}
\label{eq:arr_mxc}
\Delta_{a,c}^s = s_a^s \cdot \left( m_{a,c}^{s,2} - m_{a,c}^{s,1}\right)\textrm{.}
\end{equation}

This way of accounting for causes of death is exactly additive. However, if any all-cause mortality rate difference is close to 0, equation~\eqref{eq:arr_sen} is vulnerable, in which case we advise deriving the sensitivity using a more robust (but less precise) approach \citep{coddecomp}

\subsection*{Rescale Arriaga results}\label{rescale-arriaga-results}

 To obtain the \emph{net} impact of $\Delta_a^s$ (or $\Delta_{a,c}^s$) on overall life expectancy change we rescale using \eqref{eq_rescale} to match the life expectancy components from equation \eqref{eq_kit_e0}.

\begin{equation}
\label{eq_rescale}
\Delta_{a,c}^{s,\text{net}} = \Delta^{e_0^s} \cdot \frac{\Delta_{a,c}^s}{\Delta^s} \quad \mathrm{,}
\end{equation}

With the composition effect from equation~\eqref{eq_kit_comp} ( $\Delta^\pi$) and \eqref{eq_rescale}, we have all elements of the proposed decomposition:

\begin{equation}
\label{eq_full}
\Delta = \Delta^\pi + \sum_s \sum_a \sum_c\Delta_{a,c}^{s,\mathrm{net}}\mathrm{.}
\end{equation}

\subsection*{Application}\label{rescale-arriaga-results}

We use data from \citet{trias2023age} on sex-, age-, education-, and cause-specific death counts for individuals aged 35 and over for the years 2016-21 in Spain, which were obtained by request from the Spanish National Statistics Institute (INE). We use all-cause mortality rates for each education- and sex-specific subpopulation from the same source. Educational attainment information is not recorded on death certificates, but INE adds this variable through multiple data linkages \citet{ine2020asignacion}, including municipal population registers (``Padrón'') and the 2011 census. We redistributed deaths with missing education codes ($<2$\% of deaths) across the four educational groups within age, sex, and cause of death classes, proportional to deaths with non-missing education. We then grouped educational attainment into three categories: Low (primary education or less), Medium (secondary education), and High (tertiary or university education). We grouped causes of death from the original ICD10 4-digit codes into 17 main causes: Infectious (ICD-10 codes: A00-B99), Neoplasms (C00-D48), Blood (D50-D89), Endocrine (E00-E90), Mental (F00-F99), Nervous (G00-G99), Circulatory (00-I99), Respiratory (J00-J99), Digestive (K00-K93), Skin (L00-L99), Musculoskeletal (M00-M99), Genitourinary (N00-N99), Congenital (Q00-Q99), Ill-defined (R00-R94), External (chapters S, T, V, W, X and Y), Other causes (chapters H, O and P), and COVID-19 (U071-U072). Moreover, to separate the pandemic period we combined years into two time periods: 2016-2019 and 2020+. Our main results refer to 2016-2019, whereas extra results for 2020-2021 are given in the materials repository.
We smoothed and graduated the original data from 5-year age groups to single ages using a generalized additive model (GAM), including a P-spline over age and population offsets \citep{mgcv2011,mgcv2017}. The GAM formula is given in equation~\eqref{gam_model}, using the Quasi-Poisson family of models to account for possible overdispersion or Negative Binomial model in cases with a more substantial overdispersion:

\begin{equation}
\label{gam_model}
\log(\mu_a) = \beta_0 + f(a) + ln(pop_a)
\textrm{.}
\end{equation}

\FloatBarrier
\section*{Results}\label{results}
We present empirical results for the decomposition of the sex difference in life expectancy in 2016-2019.
Table 1 presents the components of the Kitagawa decomposition for males and females. In this context,  $\pi$ c represents the educational structure, while $e(35)$ denotes life expectancy at age 35 (``rate'' element in original Kitagawa terms). The table also gives the corresponding education structure and life expectancy differences between sexes. The final element,  $\Delta^{(e_{35}^s)}$, is the decomposition result, which gives the education group contribution.
The largest life expectancy gap is observed among individuals with Primary and Secondary education levels, and this gap is lower for those with Higher education.

\begin{table}[ht]
\centering
\begin{tabular}{l|lccccccc}
  \hline
  Education  & $\pi^f$ & $\pi^m$ & $\pi^f - \pi^m$ & $e(35)^f$ & $e(35)^m$ & $\overline{e(35)}$ & \makecell{$e(35)^f -$\\$ e(35)^m$} & $\Delta^{(e_{35}^s)}$ \\ 
  \hline
   Primary  & 0.09 & 0.10 & -0.01 & 48.21 & 42.47 & 45.34 & 5.74 & 0.54\\   
   Secondary  & 0.41 & 0.52 & -0.11 & 50.29 & 44.52 & 47.40 & 5.78 & 2.69\\ 
   Higher  & 0.50 & 0.38 & 0.12 & 51.30 & 46.92 & 49.11 & 4.38 & 1.92\\   
   \hline \hline
\end{tabular}
\caption{Elements of the Kitagawa decomposition, sex gaps in life expectancy in Spain, 2016-19}
\label{tab:my_table}
\end{table}

The corresponding difference in education-specific life expectancy is also illustrated in Figure 1. This figure additionally shows the differences observed for standard aggregated (non-stationary) versus radix-weighted life expectancy (stationary).

\begin{figure}[ht!]
    \centering
    \includegraphics[width=.8\linewidth]{manuscript/fig1.pdf}
    \caption{The Female-Male difference in $e(35)$ by education and population type, Spain 2016-19. }
    \label{fig:enter-label}
\end{figure}
Figure 2 is complementary to Figure 1 and represents the education contribution to the $e(35)$ sex gap with the educational composition component contribution, about 3 months, being added to the figure.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=.8\linewidth]{manuscript/fig2.pdf}
    \caption{Education structure contribution to total $e(35)$ sex gap, Spain 2016-19}
    \label{fig:enter-label}
\end{figure}

Figure 3 illustrates the cause-specific contributions to the life expectancy gap. The majority of this gap is explained by differences in mortality from cancer, circulatory diseases, respiratory conditions, and external causes. Neoplasms alone contribute to almost two full years of life expectancy. The only categories where males have a slight advantage in mortality are musculoskeletal and other minor causes.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=.8\linewidth]{manuscript/fig3.pdf}
    \caption{Cause-specific contribution to sex difference in total $e(35)$, Spain 2016-19. The Education structure contribution is presented in a different colour}
    \label{fig:enter-label}
\end{figure}

Further decomposition of cause-specific differences for age and education components of composed decomposition is shown in Figure 4. The majority of the difference in neoplasm mortality is concentrated in the older age group of 70-80 years. The circulatory component is observed in slightly younger ages and is more uniformly distributed between ages 60 and 80. In contrast, the respiratory component is more pronounced around age 85. The external causes of death primarily contribute to the difference in younger ages and since our data is bounded below by the age of 35, the full effect cannot be fully observed here.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=0.8\linewidth]{manuscript/fig4.pdf}
    \caption{The contributions of age, education, and major causes of death to the sex gap in total $e(35)$.}
    \label{fig:enter-label}
\end{figure}

\FloatBarrier

\section*{Discussion}\label{discussion}

In this paper, we propose a method for decomposing the differences in life expectancy that accounts for the compositional effects of education and causes of death between two populations mixed in the radix. Deriving an analytical solution for decomposition offers several advantages over alternative methods, particularly in terms of simplicity, speed, and repeatability. Unlike the linear integral method \citet{horiuchi2008decomposition}, which requires programming expertise, our approach can be executed swiftly and efficiently, even within spreadsheet-like environments. The computational efficiency of our method allows for the calculation of confidence intervals using bootstrapping.

Using the Kitagawa decomposition approach \citep{kitagawa1955components} to weigh together group-specific paired decompositions, our method can be generalized to address many decomposition problems involving structural compositional components, and it is not bound to mortality analyses. However, this post-weighting approach is most suitable for cases where populations are blended in a radix. In contrast, other approaches in the literature  \citep{torres2019contribution,su2023educational,su2024subnational,hendi2021immigration} decompose referring to the standard \emph{current rates} \citep{vaupel2002life} method of calculating life expectancy. These approaches treat the age pattern of group prevalence differently, in essence fixing group prevalence (weights) rather than making prevalence depend on mortality. We point out that mortality is often one of the major drivers of how group prevalence weights change over age.

 When incorporating information on causes of death, a potential vulnerability of our method arises when the difference in mortality rates in the denominator of equation~\eqref{eq:arr_sen} approaches zero (compare with Box 4.3 of \citet{preston2000demography}), rendering an implausible result. In this case, one may prefer to use a direct approximation or numerical estimate of the life expectancy sensitivity in equation~\eqref{eq:arr_mxc}.
